BOAST: a Metaprogramming Framework to Produce Portable and Efficient Computing Kernels for HPC Applications

### 2014-09-08 Paper/BOAST.tex
Improved Paper

\lstdefinestyle{BOAST}{
  language=Ruby, 
  basicstyle=\scriptsize, 
  backgroundcolor=\color{white}, 
  frame=single, 
  captionpos=b,
  morekeywords={BOAST, print, decl, close}
}

@@

In order to clearly differentiate what is going to be generated from what is
related to manipulations in the host language three keywords were defined. They
are \emph{print}, \emph{decl} and \emph{close}. As the language is an EDSL,
these three keywords are methods in the BOAST namespace.

@@

The \emph{decl} method is used for two purposes. First, it can be used to
declares a variable. Second, it can be used to print the beginning of a control
structure without an associated code block.

The \emph{close} method is the counterpart to the \emph{decl} method. It is used
to close a control structure without an associated code block.

@@

BOAST defines several class that are used to represent the structure of the
code. These classes can be sorted in two groups:

@@

\begin{figure}
\lstset{style=BOAST,
        caption={BOAST code}, 
        label={lst:BOAST_algebra}
}

\begin{lstlisting}
i = BOAST::Int("i") # or BOAST::Variable::new(Int, "i")
k = BOAST::Int("k", :size => 8)
l = BOAST::Real("l", :dim => [ BOAST::Dim(-5, 20) ], :local => true )
BOAST::decl i, k, l
BOAST::print i === 5
j = i + 5
BOAST::print k === j * 2
BOAST::print l[k] === 1.0
\end{lstlisting}

@@

\begin{lstlisting}
integer(kind=4) :: i
integer(kind=8) :: k
real(kind=8), dimension(-5:20) :: l
i = 5
k = (i + 5) * (2)
l(k) = 1.0_wp
\end{lstlisting}
\end{minipage}
\hspace{0.04\linewidth}
\begin{minipage}[b]{0.44\linewidth}
\centering
\lstset{style=customc,
        caption={C output.},
        label={lst:BOAST_algebra_C} }

\begin{lstlisting}
int32_t i;
int64_t k;
double l[26];
i = 5;
k = (i + 5) * (2);
l[k - (-5)] = 1.0;
\end{lstlisting}
\end{minipage}
\caption{BOAST Code Snippet for Variables and Expressions}
\label{fig:BOAST_algebra}
\end{figure}
The first and most fundamental abstraction is named Variable. Variables have a
type, a name and a set of named attributes. The existing attributes are mainly
inspired from FORTRAN. Those attributes are not limited and can be arbitrarily
enriched, allowing a lot of flexibility in Variable management.

The second abstraction is named Expression. It is used to combine variables into
algebraic or logic expressions. Most of the classical operators are overloaded
for those two abstractions and thus the syntax of the expressions are rather
straightforward. The exception is the assignment operator as it is important to
differentiate between assigning an Expression or a Variable in the ruby context
and the assignment operator in the context of a BOAST Expression. Thus the
assignment in a BOAST Expression is represented as the \emph{===} operator,
while the classical assignment is kept as the \emph{=} operator.

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{print} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed by
\emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that when
run produces the FORTRAN (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output.

@@

  If

  For

  While

  Case

@@

  Function Call

>>>>

\lstdefinestyle{BOAST}{
  language=Ruby, 
  basicstyle=\scriptsize, 
  backgroundcolor=\color{white}, 
  frame=single, 
  captionpos=b,
  morekeywords={BOAST, print, decl, open, close}
}

@@

In order to clearly differentiate what is going to be generated from what is
related to manipulations in the host language four keywords were defined. They
are \emph{print}, \emph{decl}, \emph{open} and \emph{close}. As the language is
an EDSL, these four keywords are methods in the BOAST namespace.

@@

The \emph{decl} method is used to declare variables or procedures and functions.

The \emph{open} method can be used to print the beginning of a control
structure without an associated code block.

The \emph{close} method is the counterpart to the \emph{open} method. It is used
to close a control structure without an associated code block.

@@

BOAST defines several class that are used to represent the structure of the
code. These classes can be sorted in two groups, algebraic related and control
flow related:

@@

\begin{figure}
\lstset{style=BOAST,
        caption={BOAST code}, 
        label={lst:BOAST_algebra} }
\begin{lstlisting}
i = BOAST::Int("i") # or BOAST::Variable::new(Int, "i")
k = BOAST::Int("k", :size => 8)
l = BOAST::Real("l", :dim => [ BOAST::Dim(-5, 20) ], :local => true )
BOAST::decl i, k, l
BOAST::print i === 5
j = i + 5
BOAST::print k === j * 2
BOAST::print l[k] === 1.0
BOAST::print l[k-1] === BOAST::sin(l[k])
\end{lstlisting}

@@

\begin{lstlisting}
integer(kind=4) :: i
integer(kind=8) :: k
real(kind=8), dimension(-5:20) :: l
i = 5
k = (i + 5) * (2)
l(k) = 1.0_wp
l(k - (1)) = sin(l(k))
\end{lstlisting}
\end{minipage}
\hspace{0.04\linewidth}
\begin{minipage}[b]{0.44\linewidth}
\centering
\lstset{style=customc,
        caption={C output.},
        label={lst:BOAST_algebra_C} }

\begin{lstlisting}
int32_t i;
int64_t k;
double l[26];
i = 5;
k = (i + 5) * (2);
l[k - (-5)] = 1.0;
l[k - (1) - (-5)] = sin(l[k - (-5)]);
\end{lstlisting}
\end{minipage}
\caption{BOAST Code Snippet for Variables and Expressions}
\label{fig:BOAST_algebra}
\end{figure}
The first and most fundamental abstraction is named Variable. Variables have a
type, a name and a set of named attributes. The existing attributes are mainly
inspired from FORTRAN. Those attributes are not limited and can be arbitrarily
enriched, allowing a lot of flexibility in Variable management.

The second abstraction is named Expression. It is used to combine variables into
algebraic or logic expressions. Most of the classical operators are overloaded
for those two abstractions and thus the syntax of the expressions are rather
straightforward. The exception is the assignment operator as it is important to
differentiate between assigning an Expression or a Variable in the ruby context
and the assignment operator in the context of a BOAST Expression. Thus the
assignment in a BOAST Expression is represented as the \emph{===} operator,
while the classical assignment is kept as the \emph{=} operator. Function calls
(FuncCall) are also abstracted and can be used in Expressions.

Figure~\ref{fig:BOAST_algebra} shows some basic usage of both Variables and
Expressions as well as the \emph{print} and \emph{decl} keywords. For clarity we
stayed out of BOAST namespace so BOAST related class and methods are prefixed
with \emph{BOAST::}. Listing~\ref{lst:BOAST_algebra} shows the BOAST code that
when run produces the FORTRAN (Listing~\ref{lst:BOAST_algebra_FORTRAN}) and C
(Listing~\ref{lst:BOAST_algebra_C}) output. On the last line, a call to the
\emph{sin} function is made through the automatic creation of a FuncCall object.
This automatic behaviors is limited to BOAST namespace, and thus, here, the
function call has to be prefixed with \emph{BOAST::}.

@@

\begin{figure}
\lstset{style=BOAST,
        caption={BOAST code}, 
        label={lst:BOAST_control} }
\begin{lstlisting}
\end{lstlisting}

\begin{minipage}[b]{0.50\linewidth}
\centering
\lstset{style=customf,
        caption={FORTRAN output.},
        label={lst:BOAST_control_FORTRAN} }

\begin{lstlisting}
\end{lstlisting}
\end{minipage}
\hspace{0.04\linewidth}
\begin{minipage}[b]{0.44\linewidth}
\centering
\lstset{style=customc,
        caption={C output.},
        label={lst:BOAST_control_C} }

\begin{lstlisting}
\end{lstlisting}
\end{minipage}
\caption{BOAST Code Snippet for Variables and Expressions}
\label{fig:BOAST_control}
\end{figure}

The classical control structures are implemented. If, For, While, Case are
abstractions in BOAST matching the behavior of corresponding control structures
in other languages. An exception is the For in BOAST more closely matches the
for in FORTRAN than the one in C.

Figure~\ref{fig:BOAST_control} show some basic usage of the control structures.

@@
